---
title: "Introduction to eva3dm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to eva3dm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(eva3dm)
library(riem)
library(terra)
```

The recommended workflow to evaluate your model results will include 4 steps:


1. Pre-processing of observations

2. Pre-processing of model output

3. Model Evaluation

4. Visualization

The exact steps needed to evaluation will be different for different models, different observation data-sets, and different variables. However, in this vignette is presented an example of evaluation of temperature at surface level (T2) from WRF-Chem for a simulation at 3-km of resolution for January 2016 using using data from METeorological Aerodrome Report (METAR).

1 - Pre-processing of observations:
We are using the r-package `riem` to download METAR data:

```{r metar}
start_date  <- "2016-01-01"
end_date    <- "2016-02-01"
sites       <- c("SBGR","SBKP","SBMT","SBSJ","SBSP","SBST","SBTA")

METAR  <- data.frame(date = seq.POSIXt(as.POSIXct(start_date), 
                                       as.POSIXct(end_date), 
                                       by = "hour"))
for(site in sites){
  cat("downloading METAR from:",site,"...\n")

  DATA <- riem_measures(station    = site,
                        date_start = start_date,
                        date_end   = end_date,
                        elev       = FALSE,
                        latlon     = FALSE)
  DATA <- data.frame(date = DATA$valid,
                     T2   = DATA$tmpf)
  names(DATA) <- c('date', site)
  METAR       <- merge(x     = METAR, 
                       y     = DATA, 
                       by    = "date", 
                       all   = T, 
                       sort  = TRUE)
}

```
Lets check if the observations are ok

```{r check, fig.width = 7, fig.height = 4}
plot(METAR$date, METAR[,2], ty = 'l',xlab = '',ylab = 'T2', main = 'METAR OBS')
head(METAR)
```
In this case the temperature units needs to be converted from Fahrenheit to Celsius and the data needs to be grouped or averaged by exact hours, in this case using the hourly function.
```{r observation, fig.width = 7, fig.height = 4}

METAR[,-1] <- 5/9 * (METAR[,-1]-32)
METAR      <- hourly(METAR)

plot(METAR$date, METAR[,2], ty = 'l',xlab = '',ylab = 'T2', main = 'METAR processed OBS')
head(METAR)

```

For the next step a site-list with latitude and longitude of each surface station is needed, a list of all METAR stations is available in eva3dm package.
```{r site-list}
site_list <- readRDS(paste0(system.file("extdata",package="eva3dm"),"/sites_METAR.Rds"))
head(site_list)
```

2 - Pre-processing of model output:
The model output are large files and do not allow the direct comparison with observations. The package provides a function that extracts time-series from geographic locations from a site-list. The function `extract_serie()` extracts the time-series from a list of files and saves the output in R file (in this example we are reading the R file generated by `extract_serie()`). An finally, the temperature extracted from the model is converted from Kelvin to Celsius.
```{r model}
## to extract time-series from WRF-Chem model
## wrf_files <- dir(pattern = "wrfout_d01")
## extract_serie(filelist = wrf_files, point = site_list, variable="T2", prefix="mmodel.d01", field="3d")
model_d01 <- readRDS(paste0(system.file("extdata",package="eva3dm"),"/model.d01.T2.Rds"))
model_d01[-1] <- model_d01[-1] - 273.15
```
The function `wrf_rast()` should be used, instead of `extract_serie()`, to extract the model output to evaluate against observations in regular grids (i.e. satellite products).

3 - Model Evaluation: 
The functions `eva()` can be used to calculate the statistical indexes from time-series for individual stations or all stations combined.
```{r evaluation}
table <- data.frame()
for(site in sites){
  table <- eva(mo = model_d01, ob = METAR, site = site, table = table)
}
table <- eva(mo = model_d01, ob = METAR, site = 'ALL', table = table)
print(table)
```
To evaluate using observations in regular grids, the function `sat()` is used instead of `eva()`.

4 - Visualization:
The function `overlay()` can be used to plot the results from `eva()` or `sat()` on a map. First the data is georeferenced with `%at%`.
```{r visualize, fig.width = 7, fig.height = 5.5}
spatial_table <- table %at% site_list
overlay(spatial_table, z = 'MB', main = 'T2 main bias (MB)',expand = 1.6,lim = 0.1)
masp <- terra::vect(paste0(system.file("extdata",package="eva3dm"),"/masp.shp"))
BR   <- terra::vect(paste0(system.file("extdata",package="eva3dm"),"/BR.shp"))
terra::lines(BR)
terra::lines(masp, col = 'gray')
legend_range(spatial_table$MB,y = table["ALL","MB"])
```
The WRF-Chem temperature in this example showed a very low MB, except for the SBSJ station (MB=-0.7), which is in the criteria for acceptable performance (|MB| < 0.5) and the criteria for evaluation for complex terrain for the SBSJ case (|MB| < 1.0). Another criteria for temperature are IAO > 0.8 and ME < 2.0 (ME < 3, for complex terrain), in general the model presented a reasonable performance for temperature. More information and references can be found on the documentation of `stat()` function.
